\Large Last Writer Slices and Communication Traps

### 2013-10-29 paper.tex
comitting the paper with some initial changes to the formalism.
Lots more to do!

\subsection{Formalism and Correctness}
\label{sec:lwssoundness}
To discuss the correctness of our \lwt-based last writer slice support, we
first develop a formal description of last writer slices.  We consider an
N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_i = \{p^{i}_{1},
p^{i}_{2}, \ldots, p^{i}_{k_{i}}\}$ is the sequence of $k_{i}$ dynamic
instructions executed by thread $i$.  We assume multi-threaded executions are
sequentially consistent (SC).  Without loss of generality, we assume all
instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
$Read$ or $Write$. We also assume $m$ is a uniquely addressed memory location.
Under SC, a point in an execution of $P$ after $e$ instructions have
executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q \in Q$ is a
$p^{i}_{j}$ from one of the $k_i$ instructions of thread $i$. We introduce
$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
in $Q$ to write $m$ and $t$ is the thread that executed it.  For a
never-written location $n$, $LWS[n] = (\emptyset,\emptyset)$ indicating it is
unwritten.

@@

\paragraph{Sequential Consistency and Data-races}

We assume data-race freedom and a memory model that provides SC under data-race
freedom (implying an SC execution).  We must make these assumptions because,
for performance reasons, we do not enforce the atomicity of \lwt updates and
their corresponding program accesses (a design choice made in prior
work~\cite{fasttrack,recon}).  This atomicity is also required for correctness.

We use the relation $p \prec q$ to denote that $p$ happens before $q$.  Under
data-race freedom, any write by thread $t$ to $m$, $p^{t}_{j} = Write(m)$,  is
ordered by program synchronization with $p^{t'}_{j'} = Op(m)$, a later access to
$m$ by any other thread $t'$ -- {\em i.e.}, $p^{t}_{j} \prec p^{t'}_{j'}$.  Our
compiler inserts an \lwt update $I(p^{t}_{j})$ into $P'$ before its write
operation, $p^{t}_{j}$, so $I(p^{t}_{j}) \prec p^{t}_{j}$.  Happens-before is
transitive, so $I(p^{t}_{j}) \prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec
p^{t'}_{j'}$, meaning $I(p^{t}_{j})$ happens before all accesses that
$p^{t}_{j}$ happens before {\em and before their respective \lwt updates},
$I(p^{t'}_{j'})$.  Symmetrically, for any access $p^{t''}_{j''}$ in another
thread $t''$ that precedes $p^{t}_{j}$, $p^{t''}_{j''} \prec p^{t}_{j}$.  Our
compiler inserts each $I(p^{t}_{j})$ immediately before each $p^{t}_{j}$, so
$I(p^{t''}_{j''}) \prec p^{t''}_{j''} \prec I(p^{t}_{j}) \prec p^{t}_{j}$,
happens-before ordering $I(p^{t}_{j})$ after all preceding accesses to $m$ in
other threads, {\em and after their repective \lwt updates}.  Thus, at all
points $Q$ in a data-race free execution, given $q_r = I(p^{t}_{j})$, $q_p =
p^{t}_{j}$, and $q_r \prec q_p$, $\forall_{ \{q_s | q_r \prec q_s \prec q_p
\}}: q_s \ne Op(m) \wedge q_s \ne I(Op(m)) $.  In other words, \lwt updates and
their accesses are atomic.

@@

\subsection{Debugging with Last Writer Slices}
\label{sec:debugging}
Prior work has demonstrated that it is useful during debugging to allow
programmers to ask ``why?'' questions~\cite{whylineicse,whylinechi} about the
provenance of program state.  To answer such questions, prior work has mostly
used dependence slices that include intra-thread
dependences~\cite{whylineicse,conseq,tipslicingsurvey}.  The overhead of
slicing dependence chains and tracking temporal sequences is often high,
relegating such techniques to working on traces~\cite{whylinechi} or debugging
executions.

@@

Computing last writer slices is fast enough to deploy and use
continuously (see Section~\ref{sec:eval}) and the \lwt can be saved with a core
dump.  Thus last writer slices can (and should) be made part of the standard
debugging information provided when a program fails.

>>>>

\subsection{\lwt Formalism and Correctness}
\label{sec:lwssoundness}
To discuss the correctness of our \lwt-based last writer slice support, we
first develop a formal description of last writer slices.  We consider an
N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_i = (p^{i}_{1},
p^{i}_{2}, \ldots, p^{i}_{k_{i}})$ is the sequence of $k_{i}$ dynamic
instructions executed by thread $i$.  Without loss of generality, we assume all
instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
A point in an execution of $P$ after $e$ instructions have
executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q \in Q$ is a
$p^{i}_{j}$ from one of the $k_i$ instructions of thread $i$.  We introduce
$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
in $Q$ to write $m$ and $t$ is the thread that executed it.  For a
never-written location $n$, $LWS[n] = (\emptyset,\emptyset)$ indicating it is
unwritten.

@@

\paragraph{Performance, Atomicity, and Sequential Consistency}

We do not enforce the atomicity of \lwt updates and their corresponding program
accesses.  This design choice, made in prior work~\cite{fasttrack,recon},
allows us to omit the synchronization operations that enforce the atomicity of
these operations.  The benefit of omitting these synchronization operations is
that overhead of our analysis is reduced. The downside is that our analysis is
vulnerable to violations of the atomicity of \lwt updates and their
corresponding accesses. Importantly, however, these violations can only occur
in programs with data-races and all programs with data-races do not necessarily
corrupt the \lwt.

We use our \lwt formalism from above to characterize how data-races can
interfere with \lwt updates.  The relation $p \prec q$ denotes that $p$ happens
before $q$.  In the absence of data-races, any write by thread $t$ to $m$,
$p^{t}_{j} = Write(m)$, is ordered by program synchronization with $p^{t'}_{j'}
= Op(m)$, a later access to $m$ by any other thread $t'$ -- {\em i.e.},
$p^{t}_{j} \prec p^{t'}_{j'}$.  Our compiler inserts an \lwt update
$I(p^{t}_{j})$ into $P'$ before its write operation, $p^{t}_{j}$, so
$I(p^{t}_{j}) \prec p^{t}_{j}$.  Happens-before is transitive, so without
data-races, $I(p^{t}_{j}) \prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec
p^{t'}_{j'}$, meaning $I(p^{t}_{j})$ happens before all accesses that
$p^{t}_{j}$ happens before {\em and before their respective \lwt updates},
$I(p^{t'}_{j'})$.  Symmetrically, for any access $p^{t''}_{j''}$ in another
thread $t''$ that precedes $p^{t}_{j}$, $p^{t''}_{j''} \prec p^{t}_{j}$.  Our
compiler inserts each $I(p^{t}_{j})$ immediately before each $p^{t}_{j}$, so
$I(p^{t''}_{j''}) \prec p^{t''}_{j''} \prec I(p^{t}_{j}) \prec p^{t}_{j}$,
happens-before ordering $I(p^{t}_{j})$ after all preceding accesses to $m$ in
other threads, {\em and after their repective \lwt updates}.  Thus, at all
points $Q$ in a data-race free execution, given $q_r = I(p^{t}_{j})$, $q_p =
p^{t}_{j}$, and $q_r \prec q_p$, $\forall_{ \{q_s | q_r \prec q_s \prec q_p
\}}: q_s \ne Op(m) \wedge q_s \ne I(Op(m)) $.  In other words, \lwt updates and
their accesses are atomic.

@@

\subsection{Debugging with Last Writer Slices}
\label{sec:debugging}
Prior work has demonstrated that it is useful during debugging to allow
programmers to ask ``why?'' questions~\cite{whylineicse,whylinechi} about the
provenance of program state.  To answer such questions, prior work has mostly
used dependence slices that include intra-thread
dependences~\cite{whylineicse,conseq,tipslicingsurvey}.  The overhead of
slicing dependence chains and tracking temporal sequences is often high,
relegating such techniques to working on traces~\cite{whylinechi} or debugging
executions run ``in the lab''.

@@

Computing last writer slices is fast enough to deploy and use continuously (see
Section~\ref{sec:eval}) and the \lwt can be saved with a core dump.  Thus, last
writer slices can be made part of the standard debugging information provided
when a program fails.
